---
title: 'PIM inhibition on formation of PMAr-RS-TAMs (RS-PMAr 5 d): initial exploratory analysis and excluding outliers'
author: "Michal Pawlak"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
theme: cosmo

---

## Reading count data

human genome version: ensembl GRCh38.95
counts obtained by: htseq-count

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      fig.width = 8,
                      fig.asp = 0.618,
                      out.width = "90%",
                      fig.align = "center",
                      message=FALSE, 
                      warning=FALSE)

suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("ggrepel"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("DT"))
suppressPackageStartupMessages(library("sva"))
suppressPackageStartupMessages(library("Biobase"))
suppressPackageStartupMessages(library("DESeq2"))
suppressPackageStartupMessages(library("pheatmap"))
suppressPackageStartupMessages(library("RColorBrewer"))
suppressPackageStartupMessages(library("reshape2"))
suppressPackageStartupMessages(library("ggpubr"))
suppressPackageStartupMessages(library("org.Hs.eg.db"))
suppressPackageStartupMessages(library("tibble"))
suppressPackageStartupMessages(library("EnhancedVolcano"))
suppressPackageStartupMessages(library("magrittr"))
suppressPackageStartupMessages(library("msigdbr"))
suppressPackageStartupMessages(library("fgsea"))
suppressPackageStartupMessages(library("stringr"))
suppressPackageStartupMessages(library("purrr"))
suppressPackageStartupMessages(library("readr"))
```

```{r read_se}
#we will read se object and subset to match experiment 2, set 3

se <- readRDS("../data/se.RDS")
se <- se[,c(37:42)]

rowData(se)$ENSEMBL <- rownames(se)
rownames(se) <- rowData(se)$gene_name

```

## Annotate samples

We are going to prepare a sample table contatining information about different conditions, cell types, replicates and known technical batch.

```{r meta}

# get clean meta data
meta <- data.frame(replicate =  c(1, 2, 3, 1, 2, 3),
                   treatment = c(rep("DMSO", 3), rep("MEN", 3)))


meta$SAMPLE_ID <- colData(se)[,1]
meta$SAMPLE_NAME <- colData(se)[,2]
rownames(meta) <- meta$SAMPLE_ID

# replace existing metadata

se@colData <- DataFrame(meta)

# print meta table

DT::datatable(
  meta,
  extensions = c("Buttons"),
  options = list(buttons = c("copy", "csv", "excel")
  ),
  rownames = FALSE
)
```

## PCA and pairwise correlation

Quality check and counts normalization.

```{r vsd_subset}

dds <- DESeqDataSet(se, design = ~treatment)

keep <- rowSums(counts(dds)) >= 10

dds <- dds[keep,]

vsd <- vst(dds)

dataPCA <-
  plotPCA(vsd,
  intgroup = c("treatment"),
  returnData = TRUE)

dataPCA$replicate <- vsd$replicate
dataPCA$treatment <- vsd$treatment


percentVar <- round(100 * attr(dataPCA, "percentVar"))
  

# PCA by treatment

ggplot(dataPCA, aes(PC1, PC2)) + 
  geom_point(aes(col = treatment), size = 4) + 
  xlab(paste0("PC1: ", percentVar[1], "%variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() + 
  scale_size_manual(values = c(5, 8, 12)) +
  scale_color_manual(values = c("orange", "red", "green", "brown", "blue", "violet", "grey")) + 
  theme_bw() +
  theme(axis.title = element_text(size = 16, face = "bold"),
  plot.title = element_text(size = 16, face = "bold"),
  legend.text =
  element_text(size = 14),
  legend.title = element_text(size = 16)) + 
  geom_text_repel(aes(label = treatment), hjust = 1, vjust = 0) + 
  ggtitle("PCA grouped by treatment") +
  labs(col = "treatment")
  
```

## Euclidean distance and spearman correlation heatmaps 

```{r heatmap_subset}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

#Eucledian distance heatmap
pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  col = colors,
  annotation_names_col = FALSE,
  labels_row = colData(vsd)$treatment,
  labels_col = colData(vsd)$treatment,
  show_colnames = FALSE,
  main = "Sample-to-sample Euclidean distance")

cor_vsd <- cor(assay(vsd), method = "spearman")
#correlation heatmap
pheatmap(
  cor_vsd,
  annotation_names_col = FALSE,
  labels_row = colData(vsd)$treatment,
  labels_col = colData(vsd)$treatment,
  show_colnames = FALSE,
  main = "Pairwise sample-to-sample spearman correlation")

saveRDS(vsd, file = "../data/vsd_mem_5d.RDS")
saveRDS(dds, file = "../data/dds_mem_5d.RDS")
```


```{r info}
sessionInfo()
```



