---
title: 'TAM data: initial exploratory analysis and excluding outliers'
author: "Michal Pawlak"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
theme: cosmo

---

## Reading count data

human genome version: ensembl GRCh38.95
counts obtained by: htseq-count

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      fig.width = 8,
                      fig.asp = 0.618,
                      out.width = "90%",
                      fig.align = "center",
                      message=FALSE, 
                      warning=FALSE)

suppressPackageStartupMessages(library("ggplot2"))
suppressPackageStartupMessages(library("ggrepel"))
suppressPackageStartupMessages(library("dplyr"))
suppressPackageStartupMessages(library("DT"))
suppressPackageStartupMessages(library("sva"))
suppressPackageStartupMessages(library("Biobase"))
suppressPackageStartupMessages(library("DESeq2"))
suppressPackageStartupMessages(library("pheatmap"))
suppressPackageStartupMessages(library("RColorBrewer"))
suppressPackageStartupMessages(library("reshape2"))
suppressPackageStartupMessages(library("ggpubr"))
suppressPackageStartupMessages(library("org.Hs.eg.db"))
suppressPackageStartupMessages(library("tibble"))
suppressPackageStartupMessages(library("EnhancedVolcano"))
suppressPackageStartupMessages(library("magrittr"))
suppressPackageStartupMessages(library("msigdbr"))
suppressPackageStartupMessages(library("fgsea"))
suppressPackageStartupMessages(library("stringr"))
suppressPackageStartupMessages(library("purrr"))
suppressPackageStartupMessages(library("readr"))
```

```{r read_se}
#we will read se object and subset to match experiment 1

se <- readRDS("../data/se.RDS")
se <- se[,c(1:18)]
se <- se[,c(1:4, 14:17, 5:13, 18)]

rowData(se)$ENSEMBL <- rownames(se)
rownames(se) <- rowData(se)$gene_name

```

## Annotate samples

We are going to prepare a sample table contatining information about different conditions, cell types, replicates and known technical batch.

```{r meta}

# get clean meta data
meta <- data.frame(replicate =  c(rep(1, 4), rep(2, 4), rep(1, 3), rep(2, 3), rep(3, 3), 4),
                          cell_type = c(rep(c("M0_9d", "M1", "M2", "M0_9d"),2), 
                                        rep(c("M0_10d", "RS-M", "RS-M"), 3), "RS-M"),
                          cell_line = rep("THP-1", 18),
                          stim_1 = rep("PMA", 18),
                          stim_2 = c(rep(c("ctrl", "LPS_IFNg", "IL4_IL13", "CSF1"), 2), 
                                     rep(c("ctrl", "L428", "L1236"), 3),       "L428"),
                          date = c(rep("17.10.18", 4), rep("14.11.18", 4), rep("13.09.18", 3),
                                   rep("21.03.19", 2), "28.02.19", rep("29.04.19", 3), "13.06.19"))

meta$cell_type_stim2 <- paste(meta$cell_type, meta$stim_2, sep = "_")
meta$SAMPLE_ID <- colData(se)[,1]
meta$SAMPLE_NAME <- colData(se)[,2]
rownames(meta) <- meta$SAMPLE_ID

# replace existing metadata

se@colData <- DataFrame(meta)

# print meta table

DT::datatable(
  meta,
  extensions = c("Buttons"),
  options = list(buttons = c("copy", "csv", "excel")
  ),
  rownames = FALSE
)
```

## Identify and remove batch effects

We will create DESeq2 object that will be further used in the analysis. We will apply variance stabilizing transformation `vst` to normalize data and we will have a look at PCA before and after batch (date of library preparation) effect removal. For the batch removal we will use `limma` package `removeBatchEffect` function. The function (in effect) fits a linear model to the data, including both batches and regular treatments, then removes the component due to the batch effects.

```{r vsd_and_batch}

dds <- DESeqDataSet(se, design = ~stim_2)

#change rownames to symbol
rownames(dds) <- rowData(se)$gene_name

# remove low-counterss
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

# perform vst normalization

# batched samples

vsd <- vst(dds)

dataPCA <-
  plotPCA(vsd,
  intgroup = c("cell_type_stim2"),
  returnData = TRUE)
  percentVar <- round(100 * attr(dataPCA, "percentVar"))
  
dataPCA$stim2 <- colData(vsd)$stim_2
dataPCA$replicate <- colData(vsd)$replicate
dataPCA$date <- colData(vsd)$date

# PCA by treatment
ggplot(dataPCA, aes(PC1, PC2)) + 
  geom_point(aes(col = cell_type_stim2), size = 4) + 
  xlab(paste0("PC1: ", percentVar[1], "%variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() + scale_size_manual(values = c(5, 8, 12)) +
  scale_color_manual(values = c("orange", "red", "green", "brown", "blue", "violet", "grey")) + 
  theme_bw() +
  theme(axis.title = element_text(size = 16, face = "bold"),
    plot.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.title = element_text(size = 16)) + 
  geom_text_repel(aes(label = group), hjust = 1, vjust = 0) + 
  ggtitle("PCA grouped by stimulation,\nbefore batch correction")
  
# PCA by experiment
ggplot(dataPCA, aes(PC1, PC2)) + 
  geom_point(aes(col = date), size = 4) + 
  xlab(paste0("PC1: ", percentVar[1], "%variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() + 
  scale_size_manual(values = c(5, 8, 12)) +
  scale_color_manual(values = c("orange", "red", "green", "brown", "blue", "violet", "grey")) + 
  theme_bw() +
  theme(axis.title = element_text(size = 16, face = "bold"),
  plot.title = element_text(size = 16, face = "bold"),
  legend.text =
  element_text(size = 14),
  legend.title = element_text(size = 16)) + 
  geom_text_repel(aes(label = group), hjust = 1, vjust = 0) + 
  ggtitle("PCA grouped by date of experiment,\nbefore batch correction")


# batch correction

vsd_batch <- vsd

assay(vsd_batch) <- limma::removeBatchEffect(assay(vsd_batch), vsd_batch$date)

dataPCA <-
  plotPCA(vsd_batch,
  intgroup = c("cell_type_stim2"),
  returnData = TRUE)
  percentVar <- round(100 * attr(dataPCA, "percentVar"))
  
dataPCA$stim2 <- colData(vsd_batch)$stim_2
dataPCA$replicate <- colData(vsd_batch)$replicate
dataPCA$date <- colData(vsd_batch)$date

# PCA grouped by treatment 

ggplot(dataPCA, aes(PC1, PC2)) + 
  geom_point(aes(col = cell_type_stim2), size = 4) + 
  xlab(paste0("PC1: ", percentVar[1], "%variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() + 
  scale_size_manual(values = c(5, 8, 12)) +
  scale_color_manual(values = c("orange", "red", "green", "brown", "blue", "violet", "grey")) + 
  theme_bw() +
  theme(axis.title = element_text(size = 16, face = "bold"),
  plot.title = element_text(size = 16, face = "bold"),
  legend.text =
  element_text(size = 14),
  legend.title = element_text(size = 16)) + 
  geom_text_repel(aes(label = group), hjust = 1, vjust = 0) + 
  ggtitle("PCA grouped by stimulation,\nafter batch correction")

# PCA grouped by experiment
  
ggplot(dataPCA, aes(PC1, PC2)) + 
  geom_point(aes(col = date), size = 4) + 
  xlab(paste0("PC1: ", percentVar[1], "%variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() + 
  scale_size_manual(values = c(5, 8, 12)) +
  scale_color_manual(values = c("orange", "red", "green", "brown", "blue", "violet", "grey")) + 
  theme_bw() +
  theme(axis.title = element_text(size = 16, face = "bold"),
  plot.title = element_text(size = 16, face = "bold"),
  legend.text =
  element_text(size = 14),
  legend.title = element_text(size = 16)) + 
  geom_text_repel(aes(label = group), hjust = 1, vjust = 0) + 
  ggtitle("PCA grouped by date of experiment,\nafter batch correction")
```


## Euclidean distance and spearman correlation heatmaps before batch correction 

```{r heatmap_before_batch}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

#Eucledian distance heatmap
pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  annotation_names_col = FALSE,
  col = colors,
  show_colnames = FALSE,
  labels_row = colData(vsd)$SAMPLE_NAME,
  main = "Sample-to-sample Euclidean distance")

cor_vsd <- cor(assay(vsd), method = "spearman")

#correlation heatmap
pheatmap(cor_vsd, 
  annotation_names_col = FALSE,
  labels_row = colData(vsd)$SAMPLE_NAME,
  show_colnames = FALSE,
  main = "Pairwise sample-to-sample spearman correlation")
```

## Euclidean distance and spearman correlation heatmaps after batch correction 

```{r heatmap after batch}
sampleDists <- dist(t(assay(vsd_batch)))

sampleDistMatrix <- as.matrix(sampleDists)

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

#Eucledian distance heatmap

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors,
         show_colnames = FALSE,
         annotation_names_col = FALSE,
         labels_row = colData(vsd_batch)$SAMPLE_NAME,
         main = "Sample-to-sample Euclidean distance")

cor_vsd <- cor(assay(vsd_batch), method = "spearman")

#correlation heatmap
pheatmap(cor_vsd, annotation_names_col = FALSE,
         labels_row = colData(vsd_batch)$SAMPLE_NAME,
         labels_col = colData(vsd_batch)$SAMPLE_NAME,
         show_colnames = FALSE,
         main = "Pairwise sample-to-sample spearman correlation")
```


## Exluding outliers

Based on the quality control metrics, we have decided to concetrate on selected samples from RS-macrophages co-cultured with L1236 and L428 line and selected corresponding control samples (have a look at the table below). We are going to build the model again only with those samples. We will normalize the counts by using `vst`.

```{r subset}
se <- se[, c(1, 9:14, 17)]

DT::datatable(
  as.data.frame(colData(se)),
  extensions = c("Buttons"),
  options = list(buttons = c("copy", "csv", "excel")),
  rownames = FALSE
)
```

```{r vsd_subset}

dds <- DESeqDataSet(se, design = ~stim_2)

keep <- rowSums(counts(dds)) >= 10

dds <- dds[keep,]

vsd <- vst(dds)

dataPCA <-
  plotPCA(vsd,
  intgroup = c("stim_2"),
  returnData = TRUE)
  percentVar <- round(100 * attr(dataPCA, "percentVar"))
  
dataPCA$stim2 <- colData(vsd)$stim_2
dataPCA$replicate <- colData(vsd)$replicate
dataPCA$date <- colData(vsd)$date

# PCA by treatment

ggplot(dataPCA, aes(PC1, PC2)) + 
  geom_point(aes(col = stim2), size = 4) + 
  xlab(paste0("PC1: ", percentVar[1], "%variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed() + 
  scale_size_manual(values = c(5, 8, 12)) +
  scale_color_manual(values = c("orange", "red", "green", "brown", "blue", "violet", "grey")) + 
  theme_bw() +
  theme(axis.title = element_text(size = 16, face = "bold"),
  plot.title = element_text(size = 16, face = "bold"),
  legend.text =
  element_text(size = 14),
  legend.title = element_text(size = 16)) + 
  geom_text_repel(aes(label = replicate), hjust = 1, vjust = 0) + 
  ggtitle("PCA grouped by stimulation") +
  labs(col = "stimulation")
  
```

## Euclidean distance and spearman correlation heatmaps 

```{r heatmap_subset}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

#Eucledian distance heatmap
pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  col = colors,
  annotation_names_col = FALSE,
  labels_row = colData(vsd)$stim_2,
  labels_col = colData(vsd)$stim_2,
  show_colnames = FALSE,
  main = "Sample-to-sample Euclidean distance")

cor_vsd <- cor(assay(vsd), method = "spearman")
#correlation heatmap
pheatmap(
  cor_vsd,
  annotation_names_col = FALSE,
  labels_row = colData(vsd)$stim_2,
  labels_col = colData(vsd)$stim_2,
  show_colnames = FALSE,
  main = "Pairwise sample-to-sample spearman correlation")

saveRDS(vsd, file = "../data/vsd.RDS")
saveRDS(dds, file = "../data/dds.RDS")
```


```{r info}
sessionInfo()
```



